---
title: "Reproducibility"
author: "Armin Rauschenberger"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{longtable}
  - \usepackage{placeins}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#utils::install.packages(c("readxl","nlme","emmeans","xtable"))
```

# Data

This chunk reads the clinical data, the laboratory data, and the randomisation list.

```{r read_data}
rm(list=ls())

#--- clinical data ---

# variable domains: https://en.wikipedia.org/wiki/SDTM
folder <- "I:/Projects folder/CCMS/Crypto-HIV/DNDi-5FC-02-CM (fed study)/4 - Data Management/7-Data transfers/Export files/24032023"
file <- dir(folder)
clindata <- list()
for(i in seq_along(file)){
  clindata[[i]] <- as.data.frame(readxl::read_xlsx(path=file.path(folder,file[i])))
  #alternative: xlsx::read.xlsx(path=file.path(folder,files[i]))
}

names(clindata) <- sapply(file,function(x) strsplit(x,split="_")[[1]][1])
names(clindata)[grepl(pattern="MH",x=file) & !grepl(pattern="V2",x=file)] <- "MH_old"
names(clindata)[grepl(pattern="MH.*V2",x=file)] <- "MH"

#create variable for randomisation identifier (RID)
for(i in seq_along(clindata)){
  ids <- strsplit(x=clindata[[i]]$USUBJID,split="/")
  clindata[[i]]$RID <- as.numeric(sapply(ids,function(x) ifelse(length(x)==1,NA,x[[2]])))
}

#--- randomisation list ---

folder <- "I:/Projects folder/CCMS/Crypto-HIV/DNDi-5FC-02-CM (fed study)/9 - Final analysis/Data"
file <- "RandomizationList.xlsx"
randata <- as.data.frame(readxl::read_xlsx(path=file.path(folder,file)))
colnames(randata)[colnames(randata)=="seqence"] <- "sequence"
randata$seq <- ifelse(randata$sequence==1,"seq 1 (AB)",ifelse(randata$sequence==2,"seq 2 (BA)",NA))
colnames(randata)[colnames(randata)=="Subject_ID"] <- "RID"

#--- laboratory data ---

folder <- "I:/Projects folder/CCMS/Crypto-HIV/DNDi-5FC-02-CM (fed study)/4 - Data Management/7-Data transfers/Import files/15032023_Pharmetheus/0131FRM18_DNDi-5FC-02-CM_PK_20230315/0131FRM18_DNDi-5FC-02-CM_PK_20230315"
file <- c("0131FRM18_5-fluorouracil_20230313.csv","0131FRM18_Flucytosine_20230314.csv")
labdata <- list()
for(i in seq_along(file)){
  labdata[[i]] <- utils::read.csv(file=file.path(folder,file[i]),header=T,sep=",",fileEncoding="latin1")
  colnames(labdata[[i]])[colnames(labdata[[i]])=="SUBJECTID"] <- "RID"
}
names(labdata) <- c("fluorouracil","flucytosine")

for(i in 1:2){
  for(k in c("seq 1 (AB)","seq 2 (BA)")){
    ids <- randata$RID[randata$seq==k]
    labdata[[i]]$seq[labdata[[i]]$RID %in% ids] <- k
  }
  labdata[[i]]$treat[labdata[[i]]$PERIOD==1 & labdata[[i]]$seq=="seq 1 (AB)"] <- "A"
  labdata[[i]]$treat[labdata[[i]]$PERIOD==1 & labdata[[i]]$seq=="seq 2 (BA)"] <- "B"
  labdata[[i]]$treat[labdata[[i]]$PERIOD==2 & labdata[[i]]$seq=="seq 1 (AB)"] <- "B"
  labdata[[i]]$treat[labdata[[i]]$PERIOD==2 & labdata[[i]]$seq=="seq 2 (BA)"] <- "A"
}

if(FALSE){
# variable search
pattern <- "SOC"
names <- sapply(clindata,colnames)
id <- which(grepl(pattern=pattern,x=tolower(unlist(names))))
unlist(names)[id]

sapply(clindata,function(x) any(grepl(pattern=pattern,tolower(unlist(x)))))
}
```


# descriptive statistics

This chunk defines functions.

```{r define_functions,results="asis"}
#--- functions for descriptive statistics ---

descnum <- function(x,digits=2){
  out <- c(mean=round(mean(x,na.rm=TRUE),digits=digits),
           median=round(median(x,na.rm=TRUE),digits=digits),
           range=paste0(round(min(x,na.rm=TRUE),digits=digits)," - ",round(max(x,na.rm=TRUE),digits=digits)),
           n=as.integer(sum(!is.na(x))))
  return(out)
}

desccat <- function(x,digits=2){
  count <- table(x)
  prop <- round(100*count/sum(count),digits=digits)
  out <- paste0(count," (",prop,"%)")
  names(out) <- names(count)
  out <- c(out,n=sum(!is.na(x)))
  return(out)
}

descPK <- function(x,digits=2){
  x <- x[!is.na(x)]
  if(any(x<=0)){stop("negative value")}
  out <- c(ari.mean=round(mean(x),digits=digits),
           ari.sd=round(sd(x),digits=digits),
           ari.cv=round(100*sd(x)/mean(x),digits=digits),
           median=round(median(x),digits=digits),
           min=round(min(x),digits=digits),
           max=round(max(x),digits=digits),
           geo.mean=round(exp(mean(log(x))),digits=digits),
           geo.sd=round(sqrt(exp(var(log(x)))),digits=digits), # not checked
           geo.cv=round(100*sqrt(exp(var(log(x)))-1),digits=digits),
           n=as.integer(length(x)))
  return(out)
}

desc4group <- function(x,group,digits=2,PK=FALSE){
  if(PK){
    list <- tapply(X=x,INDEX=group,FUN=descPK)
    list$both <- descPK(x=x[!is.na(group)])
  } else if(is.numeric(x)){
    list <- tapply(X=x,INDEX=group,FUN=descnum)
    list$both <- descnum(x=x[!is.na(group)],digits=digits)
  } else {
    x <- as.factor(x)
    list <- tapply(X=x,INDEX=group,FUN=desccat)
    list$both <- desccat(x=x[!is.na(group)],digits=digits)
  }
  table <- do.call(what="cbind",args=list)
  return(table)
}

list2table <- function(x){
  if(is.null(names(x))){stop("missing slot names")}
  x <- lapply(x,function(x) if(is.vector(x)){t(as.matrix(x))}else{x})
  ncol <- sapply(list,ncol)
  if(length(unique(ncol))!=1){stop("different column numbers")}
  names <- t(sapply(list,colnames))
  if(nrow(unique(names))!=1){stop("different column names")}
  for(i in seq_along(x)){
    x[[i]] <- cbind(c(names(x)[i],rep("",times=nrow(x[[i]])-1)),rownames(x[[i]]),x[[i]])
    rownames(x[[i]]) <- NULL
  }
  table <- do.call(what="rbind",args=x)
  colnames(table)[1:2] <- " "
  return(table)
}

mytable <- function(x,caption=NULL,long=FALSE){
  xtable <- xtable::xtable(x=x,caption=caption)
  hline.after <- unique(c(which(x[,1]!="")-1,nrow(x)))
  xtable::print.xtable(x=xtable,comment=FALSE,caption.placement="top",include.rownames=FALSE,hline.after=hline.after,table.placement="!h")
}

mylist <- function(x,digits=2,id="RID",align=NULL,caption=""){
  for(i in seq_along(rev(id))){
     x <- x[order(x[[rev(id)[i]]]),]
  }
  #x <- x[order(x[[id]]),]
  quote <- c(FALSE,diff(as.numeric(x[[id[1]]]))==0)
  hline <- c(diff(as.numeric(x[[id[1]]]))!=0,TRUE)
  x[[id[1]]][quote] <- "\""
  xtable <- xtable::xtable(x=x,digits=digits,caption=caption)
  if(!is.null(align)){
    xtable::align(xtable) <- align
  }
  xtable::print.xtable(xtable,include.rownames=FALSE,comment=FALSE, tabular.environment=ifelse(is.null(align),"tabular","longtable"),floating=is.null(align),caption.placement="top",hline.after=c(0,which(hline)),table.placement="!h")
}

mylog <- function(x,base=exp(1),cond=TRUE){
  if(cond){
    x <- base::log(x,base=base)
  } else {
    x
  }
  return(x)
}

#--- functions for estimating pharmacokinetic parameters---

auc <- function(conc,time){
  n <- length(time)-1
  part <- rep(NA,times=n)
  for(i in seq_len(n)){
    # method "lin up, log down"
    if(conc[i+1]>=conc[i]){
      part[i] <- (0.5*(conc[i+1]-conc[i])+conc[i])*(time[i+1]-time[i])
    } else {
      part[i] <- (time[i+1]-time[i])*(conc[i]-conc[i+1])/log(conc[i]/conc[i+1])
    }
  }
  return(sum(part))
}

cmax <- function(conc,time){
  return(max(conc))
}

tmax <- function(conc,time){
  return(time[which.max(conc)])
}

lambdaZ <- function(conc,time){
  tmax <- tmax(conc,time)
  cond <- conc>0 & time>tmax
  iter <- sum(cond)-2
  lm <- list()
  for(i in seq_len(iter)){
    lm[[i]] <- stats::lm(log(conc[cond])~time[cond])
    cond[which.max(cond)] <- FALSE
  }
  adjr2 <- sapply(lm,function(x) summary(x)$adj.r.squared)
  which.max <- which.max(adjr2)
  # prefer longer slope even if slightly lower R2
  cond <- (adjr2+1e-04)>adjr2[which.max] & seq_along(adjr2) < which.max
  if(any(cond)){
    which.max <- which(cond)
  }
  slope <- -lm[[which.max]]$coefficients["time[cond]"]
  return(slope)
}

aucInf <- function(conc,time){
  auc <- auc(conc=conc,time=time)
  lambdaZ <- lambdaZ(conc=conc,time=time)
  out <- auc + rev(conc[conc!=0])[1]/lambdaZ
  return(out)
}

atehl <- function(conc,time){
  lambdaZ <- lambdaZ(conc=conc,time=time)
  out <- log(2)/lambdaZ
  return(out)
}

fluct <- function(conc,time){
  (max(conc)-min(conc))/mean(conc)
}

#--- functions for mixed models ---

names2table <- function(x,ncol=0){
  # add rownames
  x <- cbind(rownames(x),x)
  rownames(x) <- NULL
  # add colnames
  x <- rbind(c(colnames(x)),x)
  colnames(x) <- NULL
  # add empty columns
  if(ncol>ncol(x)){
    extra <- matrix("",nrow=nrow(x),ncol=ncol-ncol(x))
    x <- cbind(x,extra)
  }
  return(x)
}

lme2latex <- function(lme,caption=NULL){
  # random effects
  random <- nlme::VarCorr(lme)[,"Variance"]
  class(random) <- "numeric"
  random <- as.matrix(round(random,digits=3))
  colnames(random) <- "variance"
  # fixed effects
  fixed <- summary(lme)$tTable[,c("Value","p-value")]
  colnames(fixed)[colnames(fixed)=="Value"] <- "estimate"
  rownames(fixed) <- gsub(pattern="TRUE",replacement="",x=rownames(fixed))
  fixed <- cbind("estimate"=format(fixed[,"estimate"],digits=1),"p-value"=format(signif(fixed[,"p-value"],digits=2)))
  # marginal means
  marginal <- matrix(data=NA,nrow=2,ncol=3,dimnames=list(c("estimate","CI"),c("FALSE","TRUE","ratio")))
  temp <- summary(emmeans::emmeans(lme,spec=c("treatment"),show.levels=TRUE),level=0.90)
  for(i in c("FALSE","TRUE")){
    marginal["estimate",i] <- round(exp(temp$emmean[temp$treatment==i]),digits=2)
    lower <- round(exp(temp$lower.CL[temp$treatment==i]),digits=2)
    upper <- round(exp(temp$upper.CL[temp$treatment==i]),digits=2)
    marginal["CI",i] <- paste0(lower,"-",upper)
  }
  # geometric mean ratio
  coefs <- summary(lme)$tTable
  ratio <- coefs["treatmentTRUE",c("Value","Std.Error","DF")]
  ci.logratio <- ratio["Value"]+c(-1,1)*stats::qt(p=0.95,df=ratio["DF"]-1)*ratio["Std.Error"]
  marginal["estimate","ratio"] <- round(exp(ratio["Value"]),digits=2)
  marginal["CI","ratio"] <- paste0(round(exp(ci.logratio),digits=2),collapse="-")
  # combine
  list <- list()
  list[[1]] <- rbind("",c("\\textbf{random effects}",rep(x="",times=3)))
  list[[2]] <- names2table(random,ncol=4)
  list[[3]] <- rbind("",c("\\textbf{fixed effects}",rep(x="",times=3)))
  list[[4]] <- names2table(fixed,ncol=4)
  list[[5]] <- rbind("",c("\\textbf{marginal means}",rep(x="",times=3)))
  list[[6]] <- names2table(marginal,ncol=4)
  table <- do.call(what="rbind",args=list)
  xtable <- xtable::xtable(table,caption=caption)
  xtable::print.xtable(xtable,sanitize.text.function=identity,comment=FALSE,include.colnames=FALSE,include.rownames=FALSE,caption.placement="top")
}

#--- find global variables in functions ---

fun <- objects(all.names=TRUE,pattern="\\.")
for(i in seq_along(fun)){
  if(fun[i]==".Random.seed"){next}
  var <- codetools::findGlobals(fun[i],merge=FALSE)$variables
  if(length(var)>0){
    stop(paste0("Global variable(s) in function ",fun[i],": ",paste0(var,collapse=", ")))
  }
}
```
 
# Screening

```{r,results="asis"}
# ISSUE: missing labels for the inclusion/exclusion criteria (reproducing listing 1 "ineligible subjects with reason for ineligibility") -> check protocol
cond <- (clindata$IE$IECAT=="INCLUSION" & clindata$IE$IESTRESC=="No") | (clindata$IE$IECAT=="EXCLUSION" & clindata$IE$IESTRESC=="Yes")
mylist(x=clindata$IE[cond,c("SUBJID","IETEST","IESTRESC")],id="SUBJID",caption="Non-included or excluded subjects.")
```

# Withdrawals

```{r,results="asis"}
list <- clindata$DS[which(clindata$DS$DSTERM=="DISCONTINUED"),]
cond <- grepl(pattern="_",x=colnames(list))
obs <- colnames(list)[cond][which(list[cond]=="Yes")]
list$obs <- paste(obs,collapse=", ")
mylist(x=list[,c("RID","obs")],caption="early withdrawals",id="RID")
```

# Analyses population

```{r,eval=FALSE}
# TASK: Reproduce table 2 and listing 3.
```

 
# Protocol deviations 
 
```{r desc_deviations,results="asis"}
#--- protocol deviations ---
list <- list()
list$any <- c(tapply(X=randata$RID,INDEX=randata$seq,FUN=function(x) sum(x %in% clindata$DV$RID)),both=sum(randata$RID %in% clindata$DV$RID))
list$major <- c(tapply(X=randata$RID,INDEX=randata$seq,FUN=function(x) sum(x %in% clindata$DV$RID[clindata$DV$DVCAT=="Major"])),both=sum(randata$RID %in% clindata$DV$RID[clindata$DV$DVCAT=="Major"]))
table <- list2table(list)
mytable(table,caption="subjects with protocol deviations")

term <- clindata$DV$DVTERM
term[grepl(pattern="DELAY.*DIFFICULTY.*BLOOD.*SAMPLING",x=term)] <- "delay blood sampling"
term[grepl(pattern="EXTRA WATER",x=term)] <- "extra water"
term[grepl(pattern="DATE OF SAMPLES.*LATE",x=term)] <- "late visit"
term[grepl(pattern="POST STUDY.*LATE",x=term)] <- "late post study"
table <- t(table(term))
mytable(table,caption="protocol deviations")

table <- clindata$DV[,c("RID","VISIT","FORM","DVTERM","DVCAT")]
table$DVTERM <- tolower(table$DVTERM)
mylist(table,align="crp{1in}p{1in}p{3in}p{0.5in}",caption="Protocol deviations.")
```

# Screening characteristics

```{r desc_demographic,results="asis"}
#--- demographic ---
numeric <- c("AGE","VSORRES_WEIGHT","VSORRES_HEIGHT","VSORRES_BMI")
for(i in seq_along(numeric)){
  class(clindata$DM[[numeric[i]]]) <- "numeric"
}
colnames(clindata$DM) <- gsub(pattern="VSORRES_",replacement="",x=colnames(clindata$DM))
ids <- strsplit(x=clindata$DM$USUBJID,split="/")
id <- as.numeric(sapply(ids,function(x) ifelse(length(x)==1,NA,x[[2]])))
group <- randata$seq[id]
list <- list()
vars <- toupper(c("age","sex","race","weight","height","BMI"))
for(i in seq_along(vars)){
  values <- clindata$DM[[vars[i]]]
  list[[vars[i]]] <- desc4group(x=values,group=group)
}
table <- list2table(list)
mytable(table,caption="demographics by sequence")
```

```{r desc_drugs,results='asis',eval=TRUE}
#--- alcohol/smoker ---
list <- list()
class(clindata$SU$SUDOSE) <- "numeric"
for(k in c("ALCOHOL","SMOKER","OTHER")){
  sub <- clindata$SU[!is.na(clindata$SU$RID) & clindata$SU$SUTRT==k,]
  group <- randata$seq[sub$RID]
  list[[k]] <- desc4group(x=sub$SUOCCUR,group=group)
  if(k!="OTHER"){
    list[[paste("IF",k)]] <- desc4group(x=sub$SUDOSE,group=group)
  }
}
table <- list2table(x=list)
mytable(table,caption="alcohol/smoker by sequence")

for(k in c("ALCOHOL","SMOKER")){
  table <- clindata$SU[!is.na(clindata$SU$RID) & clindata$SU$SUTRT==k & !is.na(clindata$SU$SUDOSE),c("RID","SUDOSE")]
  mylist(table,caption=paste("listing",k))
}
```


```{r desc_medicalhist,results="asis"}
# ISSUE: For reproducing the listing of medical history (listing 7), I can find the variable "verbatim for disease/procedure" but neither "SOC" nor "preferred term".
table <- clindata$MH[!is.na(clindata$MH$RID) & !is.na(clindata$MH$MHTERM),c("RID","MHTERM","MHSTDAT","MHENDAT","MHONGO")]
mylist(table,caption="listing medical history")

cond <- !is.na(clindata$MH$RID) & !is.na(clindata$MH$FEMALE)
list <- clindata$MH[cond,c("RID","FEMALESTATUS","MHTERM")]
list$MHTERM[list$MHTERM %in% c("SINUSITIS","HEADACHE","HEADACHES")] <- NA
list <- unique(list)
mylist(x=list,caption="contraceptive methods")
```


```{r desc_priormed,results="asis"}
mylist(x=clindata$PRIORMED[!is.na(clindata$PRIORMED$RID),c("RID","CMTRT","CMSTDAT","CMENDAT","CMONGO","CMDOSFRQ","CMINDC")],caption="Listing of prior medication")
```


```{r desc_vitalsigns,results="asis"}
class(clindata$VS$VSORRES) <- "numeric"
list <- list()
for(k in c("Temperature")){
  sub <- clindata$VS[clindata$VS$VISIT=="Screening Visit" & clindata$VS$VSTEST==k,]
  group <- randata$seq[sub$RID]
  list[[paste(k,"- norm")]] <- desc4group(x=sub$VSSTRESC,group=group)
  list[[k]] <- desc4group(x=sub$VSORRES,group=group)
}
for(k in c("Supine","Standing")){
  #sub <-  clindata$VS[clindata$VS$VISIT=="Screening Visit" & clindata$VS$VSPOS==k,]
  #table(sub$SUBJID,sub$VSSTRESC)
  for(l in c("Systolic Blood Pressure","Diastolic Blood Pressure","Pulse Rate")){
    sub <-  clindata$VS[clindata$VS$VISIT=="Screening Visit" & clindata$VS$VSPOS==k & clindata$VS$VSTEST==l,]
    group <- randata$seq[sub$RID]
    if(l=="Systolic Blood Pressure"){
      list[[paste0(k,"-norm")]] <- desc4group(x=sub$VSSTRESC,group=group)
    }
    list[[paste(k,"-",l)]] <- desc4group(x=sub$VSORRES,group=group)
  }
}
table <- list2table(list)
mytable(table,caption="vital signs by sequence")
```


```{r desc_leadecg,results="asis"}
clindata$EG$EGTEST <- gsub(pattern=", Aggregate",x=clindata$EG$EGTEST,replacement="")
class(clindata$EG$EGORRES) <- "numeric"
vars <- unique(clindata$EG$EGTEST)
list <- list()
for(i in seq_along(vars)){
  sub <- clindata$EG[clindata$EG$VISIT=="SCREENING" & clindata$EG$EGTEST==vars[i],]
  if(any(duplicated(sub$EG$RID))){stop("duplicate")}
  group <- randata$seq[sub$RID]
  if(i==1){
    list[["norm"]] <- desc4group(x=sub$EGSTRESC1,group=group)
  }
  list[[vars[i]]] <- desc4group(x=sub$EGORRES,group=group)
}
table <- list2table(list)
mytable(table)
```


```{r hematology,results="asis",eval=FALSE}
# ISSUE: variables not found for screening visit (reproducing Table 12 "hematology" and others) => These laboratory data can be found in the folder transfer datam subfolder import data. In the phase II clinical trial, however, these data will be part of the clinical data.)

clindata$LB$LBCO <- gsub(pattern=" ?-? REPE?AT ?TO?P? CONFIRM AND FOLLO?W TREND ?- CUU?RR?ENT?L?T?Y NCS|FOLLOWED-UP AT POST-STUDY VISIT|\\.",replacement="",x=clindata$LB$LBCO)
unique(clindata$LB$LBCO)
vars <- "UWBC" #c("","HEAMOGLOBIN")
for(i in seq_along(vars)){
  sub <- clindata$LB[clindata$LB$VISIT=="Screening Visit" & clindata$LB$LBCO==vars[i],]
}
```


```{r desc_bloodpressure,results="asis"}
clindata$VS$period[clindata$VS$VISIT=="Treatment Period 1: 30 hrs PD"] <- 1
clindata$VS$period[clindata$VS$VISIT=="Treatment Period 2: 30 hrs PD"] <- 2
clindata$VS$seq[clindata$VS$RID %in% randata$RID[randata$seq=="seq 1 (AB)"]] <- "seq 1 (AB)"
clindata$VS$seq[clindata$VS$RID %in% randata$RID[randata$seq=="seq 2 (BA)"]] <- "seq 2 (BA)"
clindata$VS$treat[clindata$VS$period==1 & clindata$VS$seq=="seq 1 (AB)"] <- "A"
clindata$VS$treat[clindata$VS$period==1 & clindata$VS$seq=="seq 2 (BA)"] <- "B"
clindata$VS$treat[clindata$VS$period==2 & clindata$VS$seq=="seq 1 (AB)"] <- "B"
clindata$VS$treat[clindata$VS$period==2 & clindata$VS$seq=="seq 2 (BA)"] <- "A"

vars <-  c("Systolic Blood Pressure","Diastolic Blood Pressure","Pulse Rate")
time <- unique(clindata$VS$FORM[!is.na(clindata$VS$FORM)])

for(i in seq_along(vars)){
  list <- list()
  for(j in seq_along(time)){
     sub <-  clindata$VS[which(clindata$VS$FORM==time[j] & clindata$VS$VSPOS=="Supine" & clindata$VS$VSTEST==vars[i]),]
     group <- sub$treat
     list[[paste0(vars[i],"-",time[j])]] <- desc4group(x=sub$VSORRES,group=group)
  }
  table <- list2table(list)
  mytable(table,caption=paste("Supine",vars[i],"by timepoint and treatment"))
}
```

This chunk prepares the flucytosine data (to be re-organised).

```{r}
data <- labdata$flucytosine
data$SAMPLETIME..HR.[data$SAMPLETIME..HR.=="Pre-dose (0)"] <- 0
data$CONCENTRATION[data$CONCENTRATION=="BLQ"] <- 0
data$SAMPLETIME..HR. <- as.numeric(data$SAMPLETIME..HR.)
data$CONCENTRATION <- as.numeric(data$CONCENTRATION)
```

# concentration per treatment

```{r meanconc,fig.pos="!h",fig.caption="Mean concentration by treatment."}
time <- unique(data$SAMPLETIME..HR.)
mu <- sd <- list(A=rep(NA,times=length(time)),B=rep(NA,times=length(time)))
for(i in seq_along(time)){
  for(j in c("A","B")){
    mu[[j]][i]<- mean(data$CONCENTRATION[data$SAMPLETIME..HR.==time[i] & data$treat==j],na.rm=TRUE)
    sd[[j]][i]<- sd(data$CONCENTRATION[data$SAMPLETIME..HR.==time[i] & data$treat==j],na.rm=TRUE)
  }
}


graphics::par(mfrow=c(1,2),mar=c(3.5,3.5,1,1))
for(cond in c(FALSE,TRUE)){
  if(cond){
    ylim <- c(-2,5)
  } else {
    ylim <- c(0,60)
  }
  graphics::plot(x=time,y=mylog(x=mu$A,cond=cond),type="b",col="blue",ylim=ylim,xlab="",ylab="")
  graphics::segments(x0=time,y0=mylog(x=mu$A-sd$A,cond=cond),y1=mylog(x=mu$A+sd$A,cond=cond),col="blue")
  graphics::lines(x=time,y=mylog(x=mu$B,cond=cond),type="b",col="red")
  graphics::segments(x0=time,y0=mylog(x=mu$B-sd$B,cond=cond),y1=mylog(x=mu$B+sd$B,cond=cond),col="red")
  graphics::title(xlab="time",ylab=paste("log"[cond],"conc"),line=2.5)
}
```

# concentration per subject

```{r plot,fig.pos="!h",fig.cap="Concentration by subject",fig.height=8,out.width="100%"}
ylim <- range(data$CONCENTRATION,na.rm=TRUE)
graphics::par(mfrow=c(4,3),mar=c(3.5,3.5,2,1))
for(i in unique(data$RID)){
  data_ind <- data[data$RID==i,]
  sequence <- randata$sequence[randata$RID==i]
  cond <- (sequence==1 & data_ind$PERIOD==1) | (sequence==2 & data_ind$PERIOD==2)
  graphics::plot(x=data_ind$SAMPLETIME..HR.[cond],y=data_ind$CONCENTRATION[cond],type="b",col="blue",main=paste0("ind",i),ylim=ylim,ylab="",xlab="")
  
  cond <- (sequence==1 & data_ind$PERIOD==2) | (sequence==2 & data_ind$PERIOD==1)
  graphics::lines(x=data_ind$SAMPLETIME..HR.[cond],y=data_ind$CONCENTRATION[cond],type="b",col="red")
  graphics::title(xlab="time",ylab="conc",line=2.5)
}
```


# compute PK parameters

```{r pharmacokinetics,results="asis"}
if(TRUE){
  # taking into account difference between planned and actual sampling time
  head(clindata$PC)
  planned <- as.POSIXct(clindata$PC$PCTPTREF,format="%H:%M")
  actual <- as.POSIXct(clindata$PC$PCDTC,format="%H:%M")
  diff <- difftime(time1=actual,time2=planned,units="hours")
  data.frame(planned,actual,diff)
  clindata$PC$period <- gsub(pattern="Treatment Period",replacement="",x=clindata$PC$VISIT)
  clindata$PC$period <- gsub(pattern=": 30 hrs PD",replacement="",x=clindata$PC$period)
  clindata$PC$period <- as.numeric(clindata$PC$period)
  clindata$PC$time <- as.numeric(ifelse(clindata$PC$PC_SAMPLING_TIME=="Pre-dose",0,ifelse(clindata$PC$PC_SAMPLING_TIME=="0,5",0.5,clindata$PC$PC_SAMPLING_TIME)))
  for(i in seq_along(diff)){
    if(diff[i]==0){
      next
    } else {
      id <- data$RID==clindata$PC$RID[i] & data$PERIOD==clindata$PC$period[i] & data$SAMPLETIME..HR.==clindata$PC$time[i]
      if(sum(id)!=1){stop("invalid")}
      data$SAMPLETIME..HR.[id] <- data$SAMPLETIME..HR.[id] + diff[i]
    }
  }
}

n <- max(data$RID)
own <- other <- list()
own[[1]] <- other[[1]] <- own[[2]] <- other[[2]] <- data.frame(cmax=rep(NA,times=n),tmax=NA,auc=NA,lambdaZ=NA,aucInf=NA,atehl=NA,fluct=NA)
for(i in seq_len(n)){
  
  for(j in 1:2){
  
  cond <- data$RID==i & data$PERIOD==j
  conc <- data$CONCENTRATION[cond]
  time <- data$SAMPLETIME..HR.[cond]
  
  #if(i==9){next} # to reproduce the official results
  if(all(is.na(conc))){next}
  if(any(diff(time)<=0)){stop()}
  if(any(conc<0)){stop()}
  
  # own functions
  
  own[[j]]$cmax[i] <- cmax(conc=conc,time=time)
  own[[j]]$tmax[i] <- tmax(conc=conc,time=time)
  own[[j]]$auc[i] <- auc(conc=conc,time=time)
  own[[j]]$lambdaZ[i] <- lambdaZ(conc=conc,time=time)
  own[[j]]$aucInf[i] <- aucInf(conc=conc,time=time)
  own[[j]]$atehl[i] <- atehl(conc=conc,time=time)
  own[[j]]$fluct[i] <- fluct(conc=conc,time=time)
  
  # other functions

  other[[j]]$cmax[i] <- max(conc)
  other[[j]]$tmax[i] <- time[which.max(conc)]
  other[[j]]$auc[i] <- NonCompart::AUC(x=time,y=conc,down="Log")[sum(cond),"AUC"]
  other[[j]]$lambdaZ[i] <- NonCompart::BestSlope(x=time,y=conc)["LAMZ"]
  other[[j]]$aucInf[i] <- other[[j]]$auc[i] + rev(conc[conc!=0])[1]/other[[j]]$lambdaZ[i]
  
  # function with all output:
  all <- NonCompart::sNCA(x=time,y=conc,down="Log")[c("CMAX","TMAX","AUCLST","LAMZ","AUCIFO","LAMZHL")]
  other[[j]]$atehl[i] <- all["LAMZHL"]
  
   # PK::auc(conc=conc,time=time,design="complete")$est
   # PKNCA::pk.calc.auc(conc=conc,time=time,auc.type="AUClast",method="lin up/log down")
   # NonCompart::AUC(x=time,y=conc,down="Log")[sum(cond),"AUC"]
  
  }
}

#round(own[[1]],digits=4)==round(other[[1]],digits=4)
#round(own[[2]],digits=4)==round(other[[2]],digits=4)

pkdata <- rbind(cbind(subject=seq_len(n),own[[1]],period=1,sequence=randata$sequence),
cbind(subject=seq_len(n),own[[2]],period=2,sequence=randata$sequence))
pkdata$treatment <- (pkdata$sequence==1 & pkdata$period==2) | (pkdata$sequence==2 & pkdata$period==1)

# summary by treatment
vars <- c("cmax","tmax","auc","aucInf","lambdaZ","atehl")
list <- sapply(vars,function(x) desc4group(x=pkdata[[x]][pkdata$subject!=9],group=pkdata$treatment[pkdata$subject!=9],PK=TRUE),simplify=FALSE)
table <- list2table(list)
mytable(table,caption="PK parameters by treatment")

#listing by subject
mylist(x=pkdata[,c("subject","treatment","period","cmax","tmax","auc","aucInf","lambdaZ","atehl")],id=c("subject","treatment"),caption="Listing of PK parameters by subject.")
```

\FloatBarrier

# mixed model

```{r mixedmodel,results="asis"}
vars <- c("cmax","auc","aucInf")
for(i in seq_along(vars)){
  lme <- nlme::lme(as.formula(paste0("log(",vars[i],")~sequence+period+treatment")),random=~1|subject,data=pkdata[!is.na(pkdata$cmax),],method="REML")
  lme2latex(lme,caption=paste0("Mixed model for log(",vars[i],"), with a random intercept on the subject level."))
}
```


<!--
# unorganised stuff for mixed models (including estimated marginal means a.k.a. 'least squares means', alternative calculation of geometric mean ratio without mixed models)

```{r,eval=FALSE}
#--- extract estimated marginal means (SAS: least squares geometric means) ---
emmeans <- emmeans::emmeans(lme,spec=c("treatment"),show.levels=TRUE)
mu <- exp(summary(emmeans)$emmean)
lower <- exp(summary(emmeans,level=0.90)$lower.CL)
upper <- exp(summary(emmeans,level=0.90)$upper.CL)

#--- reproducing this manually ---
exp(mean(predict(object=lme,newdata=expand.grid(subject=NA,period=c(1,2),sequence=c(1,2),treatment=FALSE),level=0)))
exp(mean(predict(object=lme,newdata=expand.grid(subject=NA,period=c(1,2),sequence=c(1,2),treatment=TRUE),level=0)))

#--- calculate geometric mean ratio and CI  ---
pre <- pkdata$cmax[!pkdata$treatment][order(pkdata$subject[!pkdata$treatment])]
post <- pkdata$cmax[pkdata$treatment][order(pkdata$subject[pkdata$treatment])]

print.ci <- function(mu,ci,confl,exp=TRUE){
  if(!exp){stop("not implemented")}
  paste0(round(exp(mu),digits=2),
         " [",
         confl,
         "% CI: ",
         round(exp(ci[1]),digits=2),
         "-",
         round(exp(ci[2]),digits=2),
         "]")
}

geo.mean.ratio <- function(x,y,paired=TRUE,confl=90){
  quantile <- 0.5 + confl/200
  if(any(x<0|y<0,na.rm=TRUE)){stop("negative")}
  if(paired){
    cond <- !is.na(x) & !is.na(y)
    x <- x[cond]
    y <- y[cond]
    n <- length(x)
    log.r <- log(y/x)
    mu <- mean(log.r)
    se <- sd(log.r)/sqrt(n)
    ci <- mu+c(-1,1)*stats::qt(p=quantile,df=n-1)*se
    print.ci(mu=mu,ci=ci,confl=confl)
  } else {
    warning("This code is unchecked!")
    x <- x[!is.na(x)]
    nx <- length(x)
    y <- y[!is.na(y)]
    ny <- length(y)
    log.x <- log(x)
    log.y <- log(y)
    mu <- mean(log.y)-mean(log.x) # =log(exp(mean(log.y))/exp(mean(log.x)))
    se <- sqrt(sd(log.x)^2/nx+sd(log.y)^2/ny)
    ci <- mu + c(-1,1)*stats::qt(p=quantile,df=min(nx-1,ny-1))*se
    print.ci(mu=mu,ci=ci,confl=confl)
  }
}

geo.mean.ratio(x=pre,y=post,paired=TRUE)
# if not matched: geo.mean.ratio(x=pre,y=post,paired=FALSE)
# NB: The result is slightly different because of adjustment for other variables.

#--- geometric mean ---

geo.mean <- function(x,confl=90){
  quantile <- 0.5 + confl/200
  x <- x[!is.na(x)]
  if(any(x<0)){stop("negative")}
  n <- length(x)
  log.x <- log(x)
  mu <- mean(log.x)
  se <- sd(log.x)/sqrt(n)
  ci <- mu+c(-1,1)*stats::qt(p=quantile,df=n-1)*se
  print.ci(mu=mu,ci=ci,confl=confl)
}

geo.mean(pre)
geo.mean(post)
# NB: The result is slightly different because of adjustment for other variables.

ari.mean <- function(x,confl=90){
  quantile <- 0.5 + confl/200
  x <- x[!is.na(x)]
  n <- length(x)
  mu <- mean(x)
  se <- sd(x)/n
  ci <- mu+c(-1,1)*stats::qt(p=quantile,df=n-1)*se
  print(ci)
}

ari.mean.ratio <- function(x,y,paired=TRUE,confl=90){
  quantile <- 0.5 + confl/200
  if(paired){
      cond <- !is.na(x) & !is.na(y)
      x <- x[cond]
      y <- y[cond]
      n <- length(x)
      r <- y/x
      mu <- mean(r)
      se <- sd(r)/sqrt(n)
      ci <- mu+c(-1,1)*stats::qt(p=quantile,df=n-1)*se
  } else {
      warning("This code is unchecked!")
      x <- x[!is.na(x)]
      nx <- length(x)
      y <- y[!is.na(y)]
      ny <- length(y)
      mu <- log(mean(y)/mean(x))
      se <- sqrt(sd(y)^2/ny+sd(x)^2/nx)
      ci <- exp(mu+c(-1,1)*stats::qt(p=quantile,df=min(nx-1,ny-1))*se)
  }
  return(ci)
}

x <- abs(rnorm(100))+rep(c(0,1),each=50)
y <- abs(rnorm(100))+rep(c(0.5,1.5),each=50)
ari.mean.ratio(x=x,y=y,paired=TRUE,confl=95)
stats::t.test(x=y/x,mu=1)

x <- abs(rnorm(100))
y <- abs(rnorm(100))
ari.mean.ratio(x=x,y=y,paired=FALSE,confl=95)
stats::t.test(x)
```
-->
